Measurement Bias from Address Aliasing

### 2014-08-23 paper.tex
Comment on other archs and ht
Explain additional bias effect for Core, and include comment on
hyper-threading. Various smaller rephrasing and fixes.

% Motivation -- Problem statement -- Approach -- Results -- Conclusions
\begin{abstract} % Should be 150 -- 200 words max
Understanding program behavior and obtaining accurate measurements is important in performance analysis.
However, recent research has shown that performance measurements can be \emph{biased} by external factors in unpredictable ways.
Seemingly irrelevant variables, such as the length of your user name, can impact program performance.
In this paper, we identify an important underlying mechanism that causes this type of measurement bias on modern Intel microarchitectures.
Our approach is to search for correlations over multiple environment configurations, using a large set of hardware performance counter measurements.
We find that \emph{address aliasing} can explain bias from two external factors: environment variable size, and characteristics of dynamically linked heap allocators.

@@

\pgfplotstableread{bin/microkernel-cycles-core2.dat}{\microkernelcoretwotable}
\begin{figure}[t]
\begin{minipage}[t]{0.40\textwidth}
  \vspace{3mm}
  \begin{lstlisting}[frame=tb]
  static int i, j, k;
  int main() {
      int g = 0, inc = 1;
      for (; g < 65536; g++) {
          i += inc;
          j += inc;
          k += inc;
      }
      return 0;
  }
  \end{lstlisting}
\end{minipage}
%\hfill
\begin{minipage}[t]{0.55\textwidth}
  \begin{flushleft}
    %\fbox{
    \begin{tikzpicture}
      \begin{axis}[
          title = Core,
          width = 7.8cm,
          height = 4.5cm,
          font = \footnotesize,
          xlabel=Bytes added to environment,
          ylabel=Cycles,
          domain = 0:4096,
          xtick = {0,1024,...,4096},
          xmin = 0,
          xmax = 4096,
          cycle list name = exotic
        ]
        \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \microkernelcoretwotable ;
      \end{axis}
    \end{tikzpicture}
    %}
  \end{flushleft}
\end{minipage}
\label{fig:motivation}
\caption{Microkernel first presented by \citeN{Mytkowicz:2009:WrongData}, showing bias to environment size. Note the clear spike around 3200 bytes, where the number of cycles executed nearly doubles. Measuring performance in this configuration would give a biased result.}
\end{figure}

@@

We also have to ensure we are not affected by measurement bias beyond what we are actually trying to observe, and in general follow best practices on gathering data~\cite{Mytkowicz:2009:WrongData}.
Most importantly, this means keeping the memory address space under control.
For security reasons, addresses of stack, heap and dynamic libraries are often randomized at load time, a technique known as \emph{Address Space Layout Randomization}~\cite{Pax:ASLR,Bhatkar:AddressObfuscation}. 
By disabling ASLR, we are able to execute the same program multiple times with identical virtual address spaces.
All experiments are done on a dedicated machine under minimal load, with \emph{frequency scaling} disabled to keep the CPU's clock speed fixed.
Finally, we disable \emph{Hyper-threading}, which reduces the possibility for resource contention between threads. Also some PMU events require HT to be turned off, or are inaccurate with HT.

@@

While optimizations such as these are good on the average case, there are corner cases. 
In particular, an event known as ``4K aliasing'' can occur when the memory addresses of a store followed by a load differ by a multiple of 4096 bytes.
A store to address 0x601020 followed by a load to address 0x821020 is an aliasing pair, because the 12-bit address suffix of 0x020 is the same in both. 
Despite being independent, in these cases the memory order subsystem generates \emph{false} dependencies, causing the load to be reissued.
The number of times this happens can be counted by the following event:
\begin{describe}{{pad}}
  \item[{\small LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS.}] \emph{``Counts the number of loads that have partial address match with preceding stores, causing the load to be reissued.''} \cite[B.3.4.4]{OptimizationManual}
\end{describe}
This event is listed in the manual for microarchitectures going back to Nehalem i7, including Ivy Bridge and Haswell CPUs~\cite{Volume3B}.
Older architectures such as Core 2 do not have this particular event listed, however 4K aliasing is covered by a more general event:
\begin{describe}{{pad}}
  \item[{\small LOAD\_BLOCKS.OVERLAP\_STORE.}] \emph{``Loads that partially overlap an earlier store, or 4-Kbyte aliased with a previous store.''} \cite[Table 19-17]{Volume3B}
\end{describe}

@@

\subsection{Microkernel Analysis}
\label{sec:microkernel}
To illustrate how address aliasing can cause bias, we revisit the example first presented in ``Producing Wrong Data Without Doing Anything Obviously Wrong!'' by \citeN{Mytkowicz:2009:WrongData}, a small C program listed in Figure~\ref{fig:motivation}. 
This example is interesting for several reasons; the bias effects are significant and easily reproducible, while the code itself is simple and straightforward to analyze.
Still, no satisfactory explanation as to what actually causes bias was given in the original paper.

@@

Measuring the performance over different environment sizes for our Haswell setup reveals a similar pattern to the one shown in Figure~\ref{fig:motivation} for Core~2.
Performance counter measurements of cycle counts are shown for 512 different environment sizes in Figure~\ref{fig:envbias}.
We measure every 16 byte increment of environment size, covering two 4K periods of initial stack addresses.
A finer sampling is not necessary, because the stack is by default aligned to 16 bytes.
The program is compiled with {\small{GCC}} using no optimization, as any optimization would likely disregard most of the function as redundant code, reducing it to return zero immediately.

@@

% Detailed comparison between median and worst cases.
\begin{table}
  \tbl{Events with significant correlation to cycle count\label{tab:loopcorrelation}}{
    \pgfplotstabletypeset[
      int detect, % Output whole numbers for counter values
      col sep=comma,
      columns={Performance counter, Median, [index]3, [index]4},
      column type=r,
      columns/Performance counter/.style={
        string type, 
        column type=l,
        column type/.add={|}{},
        postproc cell content/.append code={
          \pgfkeysalso{@cell content=\perfctr{##1}}
        }
      },
      every head row/.style={
        output empty row,
        before row={\hline
          Performance counter & Median & Spike 1 & Spike 2 \\
        },
        after row=\hline\hline
      },
      every last row/.style={after row=\hline},
      every last column/.style={column type/.add={}{|}}
    ]{bin/microkernel-comparison-haswell.csv}
  }
  \begin{tabnote}
    \Note{Note:}{Some correlating performance events are omitted, for example bus-cycles which will naturally vary with total cycle count.}
  \end{tabnote}
\end{table}

We see the most extreme change from median to worst case in the number of alias events. 
If we plot the graph for address aliasing, we see that it is near zero everywhere and spikes at exactly the points we observe bias. 
The results show a high number of resource stalls and pending memory loads when the spikes occur, which is consistent with address aliasing issues.
On the other end we get a much lower number of reservation station (RS) stalls in the aliasing case, with a reduction from around 272,000 to 136,000.
The reservation station buffers micro-ops for scheduling to the execution units, and a stall event means that there are no free slots available~\cite[Table 19-2]{Volume3B}.
Fewer stalls in the aliasing case could indicate less contention on the reservation station.
This probably has to do with the overall \emph{decrease} in the number of micro-ops executed per port, as all but one of the in total 8 execution ports have fewer operations going in parallel.
Note that the number of micro-ops \emph{retired} overall does not change.
The memory disambiguation system automatically assumes dependency between a load and a previous store that are aliased, which can limit the potential for issuing many instructions simultaneously and out of order~\cite[Page 2-20]{OptimizationManual}.
In our case, lower occupation of execution ports and less pressure on the reservation station can be explained by the CPU not being able to issue more operations while waiting for an aliased memory access to be resolved.

@@

\subsection{Other Architectures}
While we show only the analysis for Haswell here, the same approach gave similar results for all CPUs we tested.
The biggest practical difference occur for Core 2, where a different set of performance events were used, in particular load\_blocks.overlap_store indicating alias.
Notably, this event was also found to be significant in the original paper by \citeN{Mytkowicz:2009:WrongData}.
We were able to verify that partial address overlap was indeed occuring synchronously with this event spiking, and we draw the conclusion that address aliasing must be the underlying explanation.
Based on our results, and existing documentation on available performance events, it is reasonable to assume that the address alias issue is present in (at least) Intel microarchitectures Core, Nehalem, Sandy Bridge and Haswell.

\subsection{Hyper-Threading}
With Hyper-Threading we see smalls spikes like Core2, likely cache bank issues. % todo: fill in something here

@@

Table \ref{tab:mallocompare} illustrates how using a different allocator can affect potential aliasing between heap pointers.\footnote{Switching allocator was done by setting the LD\_PRELOAD environment variable.}
We observe the addresses of two equally sized \texttt{char} buffers allocated with \texttt{malloc} for different size parameters.
Equal three digit address suffix indicate an aliasing pair.
In addition to glibc, where the heap allocator is called \emph{ptmalloc}, we look at the following alternatives:
\begin{enumerate}
  \item Thread-Caching Malloc (tcmalloc) by Google~\cite{TCMalloc}.
  \item jemalloc, originally developed for FreeBSD~\cite{JEMalloc}.
  \item Hoard~\cite{Berger:2000:Hoard}.
\end{enumerate}
All of the above focus heavily on performance in multithreaded environments.
Heap allocation is intrinsically inefficient in that all threads share the same address space, leading to a high potential lock contention on memory accesses.
Here, we only look at behavior for a single thread, and whether the addresses returned alias or not.
\begin{table}[t]
  \label{tab:mallocompare}
  \tbl{Addresses returned by different heap allocators when allocating pairs of equally sized buffers.}{
    \pgfplotstabletypeset[
      col sep=comma,
      string type,
      columns={Allocation, 64, 5120, 1048576},
      column type=r,
      columns/Allocation/.style={
        string type, 
        column type=l,
        column type/.add={|}{}
      },
      every head row/.style={
        output empty row,
        before row={\hline
           & 64 B & 5,120 B & 1,048,576 B \\
          },
        after row=\hline\hline,
      },
      every last row/.style={after row=\hline},
      every last column/.style={column type/.add={}{|}},
      every odd row/.style={after row=\hline},
    ]{bin/malloc-comparison.csv}
  }
  \begin{tabnote}
    \Note{Note:}{Memory mapped addresses starting with the 0x2aa... prefix is an artifact of using \emph{make} to generate results. Executing the test program from \emph{bash} directly result in 0x7fff... prefixes.
    The difference is not important to our discussion.}
  \end{tabnote}
\end{table}

>>>>

% Motivation -- Problem statement -- Approach -- Results -- Conclusions
\begin{abstract} % Should be 150 -- 200 words max
Understanding program behavior and obtaining accurate measurements is important in performance analysis.
However, recent research has shown that performance measurements can be \emph{biased} by external factors in unpredictable ways.
Seemingly irrelevant variables, such as the length of your user name, can impact program performance.
In this paper, we identify an important underlying mechanism that causes this type of measurement bias on modern Intel microarchitectures.
Our approach is to search for correlations over multiple environment configurations, using a large set of hardware performance counter measurements.
We find that \emph{address aliasing} can explain bias from two external factors; environment variable size, and characteristics of dynamically linked heap allocators.

@@

\pgfplotstableread{bin/microkernel-cycles-core2.dat}{\microkernelcoretwotable}
\begin{figure}[t]
\begin{minipage}[t]{0.40\textwidth}
  \vspace{3mm}
  \begin{lstlisting}[frame=tb]
  static int i, j, k;
  int main() {
      int g = 0, inc = 1;
      for (; g < 65536; g++) {
          i += inc;
          j += inc;
          k += inc;
      }
      return 0;
  }
  \end{lstlisting}
\end{minipage}
%\hfill
\begin{minipage}[t]{0.55\textwidth}
  \begin{flushleft}
    %\fbox{
    \begin{tikzpicture}
      \begin{axis}[
          title = Core,
          width = 7.8cm,
          height = 4.5cm,
          font = \footnotesize,
          xlabel=Bytes added to environment,
          ylabel=Cycles,
          domain = 0:4096,
          xtick = {0,1024,...,4096},
          xmin = 0,
          xmax = 4096,
          cycle list name = exotic
        ]
        \addplot[ycomb] table[x expr = \thisrowno{0}*16, y = cycles:u] \microkernelcoretwotable ;
      \end{axis}
    \end{tikzpicture}
    %}
  \end{flushleft}
\end{minipage}
\label{fig:motivation}
\caption{Microkernel first presented by \citeN{Mytkowicz:2009:WrongData}, showing bias to environment size. Note the clear spike at around 3200 bytes, where the number of cycles executed nearly doubles. Measuring performance in this configuration would give a biased result.}
\end{figure}

@@

We also have to ensure we are not affected by measurement bias beyond what we are actually trying to observe, and in general follow best practices on gathering data~\cite{Mytkowicz:2009:WrongData}.
Most importantly, this means keeping the memory address space under control.
For security reasons, addresses of stack, heap and dynamic libraries are often randomized at load time, a technique known as \emph{Address Space Layout Randomization}~\cite{Pax:ASLR,Bhatkar:AddressObfuscation}. 
By disabling ASLR, we are able to execute the same program multiple times with identical virtual address spaces.
All experiments are done on a dedicated machine under minimal load, with \emph{frequency scaling} disabled to keep the CPU's clock speed fixed.
Finally, we disable \emph{Hyper-threading} (HT), which reduces the possibility for resource contention between threads.
Some PMU events also require HT to be turned off, or produce inaccurate results with HT enabled.

@@

While optimizations such as these are good on the average case, there are corner cases. 
In particular, an event known as ``4K aliasing'' can occur when the memory addresses of a store followed by a load differ by a multiple of 4096 bytes.
A store to address 0x601020 followed by a load to address 0x821020 is an aliasing pair, because the 12-bit address suffix of 0x020 is the same in both. 
Despite being independent, in these cases the memory order subsystem generates \emph{false} dependencies, causing the load to be reissued.
The number of times this happens can be counted by the following event:
\begin{describe}{{pad}}
  \item[{\small LD\_BLOCKS\_PARTIAL.ADDRESS\_ALIAS.}] \emph{``Counts the number of loads that have partial address match with preceding stores, causing the load to be reissued.''} \cite[B.3.4.4]{OptimizationManual}
\end{describe}
This event is listed in the manual for microarchitectures going back to Nehalem i7, including Ivy Bridge and Haswell CPUs~\cite{Volume3B}.
Older architectures such as Core do not have this particular event listed, however 4K aliasing is covered by a more general event:
\begin{describe}{{pad}}
  \item[{\small LOAD\_BLOCKS.OVERLAP\_STORE.}] \emph{``Loads that partially overlap an earlier store, or 4-Kbyte aliased with a previous store.''} \cite[Table 19-17]{Volume3B}
\end{describe}

@@

\subsection{Microkernel Analysis}
\label{sec:microkernel}
To illustrate how address aliasing can cause bias, we revisit the example first presented in ``Producing Wrong Data Without Doing Anything Obviously Wrong!'' by \citeN{Mytkowicz:2009:WrongData}, a small C program reproduced here in Figure~\ref{fig:motivation}.
This example is interesting for several reasons; the bias effects are significant and easily reproducible, while the code itself is simple and straightforward to analyze.
Still, no satisfactory explanation as to what actually causes bias was given in the original paper.
The following outlines our data and analysis of this program conducted on the ``Haswell'' configuration.
Similar experiments were performed on each of the other experimental setups as well, all arriving at the same conclusions.

@@

Performance counter measurements of cycle counts are shown for 512 different environment sizes in Figure~\ref{fig:envbias}.
We measure every 16 byte increment of environment size, covering two 4K periods of initial stack addresses.
A finer sampling is not necessary, because the stack is by default aligned to 16 bytes.
The program is compiled with {\small{GCC}} using no optimization, as any optimization would likely disregard most of the function as redundant code, reducing it to return zero immediately.

@@

% Detailed comparison between median and worst cases.
\begin{table}
  \tbl{Events with significant correlation to cycle count.\label{tab:loopcorrelation}}{
    \pgfplotstabletypeset[
      int detect, % Output whole numbers for counter values
      col sep=comma,
      columns={Performance counter, Median, [index]3, [index]4},
      column type=r,
      columns/Performance counter/.style={
        string type, 
        column type=l,
        column type/.add={|}{},
        postproc cell content/.append code={
          \pgfkeysalso{@cell content=\perfctr{##1}}
        }
      },
      every head row/.style={
        output empty row,
        before row={\hline
          Performance counter & Median & Spike 1 & Spike 2 \\
        },
        after row=\hline\hline
      },
      every last row/.style={after row=\hline},
      every last column/.style={column type/.add={}{|}}
    ]{bin/microkernel-comparison-haswell.csv}
  }
  \begin{tabnote}
    \Note{Note:}{Some correlating performance events are omitted, for example bus-cycles which will naturally vary with total cycle count.}
  \end{tabnote}
\end{table}

We see the most extreme change from median to worst case in the number of alias events. 
If we plot the graph for address aliasing, we see that it is near zero everywhere and spikes at exactly the points we observe bias. 
The results show a high number of resource stalls and pending memory loads when the spikes occur, which is consistent with address aliasing issues.
On the other end we get a much lower number of reservation station (RS) stalls in the aliasing case, with a reduction from around 272,000 to 136,000.
The reservation station buffers micro-ops for scheduling to the execution units, and a stall event means that there are no free slots available~\cite[Table 19-2]{Volume3B}.
Fewer stalls in the aliasing case could indicate less contention on the reservation station.
This probably has to do with the overall \emph{decrease} in the number of micro-ops executed per port, as all but one of the in total 8 execution ports have fewer operations going in parallel.
Note that the number of micro-ops \emph{retired} overall does not change.
The memory disambiguation system automatically assumes dependency between a load and a previous store that are aliased, which can limit the potential for issuing many instructions simultaneously and out of order~\cite[Page 2-20]{OptimizationManual}.
In our case, lower occupation of execution ports, and less pressure on the reservation station, can be explained by the CPU not being able to issue more operations while waiting for an aliased memory access to be resolved.

@@

\subsection{Other Architectures}
While we show only the analysis for Haswell here, the same approach gave similar results for all CPUs we tested.
The biggest practical difference occur for Core 2, where a significantly different set of performance events were used.
In particular; load\_blocks.overlap_store indicating alias.
Notably, this event was also found to be significant in the original paper by \citeN{Mytkowicz:2009:WrongData}.
We were able to verify that partial address overlap was indeed occuring at the same time this event is spiking, and we draw the conclusion that address aliasing must be the underlying explanation.
However, this only explains part of what is going on in the Core setup. 
As shown in Figure~\ref{fig:motivation}, there seem to be yet another bias effect visible at every fourth environment increment.
These smaller spikes correlate with the performance event \emph{load\_blocks.std}, which is described as ``loads blocked by a preceding store with unknown data''~\cite[Table 19-17]{Volume3B}.
Our assumption is that this is a different phenomenon, not related to 4K aliasing.
We were only able to observe the effect on this particular architecture, and did not investigate the root cause further.

Based on our results, and existing documentation on available performance events, it is reasonable to assume that the address alias issue is present in (at least) Intel microarchitectures Core, Nehalem, Sandy Bridge and Haswell.

\subsection{Hyper-Threading}
As described in Section~\ref{sec:methodology}, disabling support for Hyper-Threading on applicable architectures was done in order to avoid any potential interference while collecting data.
However, this begs the question if doing so might affect the existence of aliasing itself.
To verify this was not the case, we ran the experiment again on the Haswell setup with Hyper-Threading enabled.
Results for cycles executed and alias events were virtually identical, indicating that configuration of HTT either way is not impactful to our conclusions.

@@

\begin{table}[t]
  \tbl{Addresses returned by different heap allocators when allocating pairs of equally sized buffers.\label{tab:mallocompare}}{
    \pgfplotstabletypeset[
      col sep=comma,
      string type,
      columns={Allocation, 64, 5120, 1048576},
      column type=r,
      columns/Allocation/.style={
        string type, 
        column type=l,
        column type/.add={|}{}
      },
      every head row/.style={
        output empty row,
        before row={\hline
           & 64 B & 5,120 B & 1,048,576 B \\
          },
        after row=\hline\hline,
      },
      every last row/.style={after row=\hline},
      every last column/.style={column type/.add={}{|}},
      every odd row/.style={after row=\hline},
    ]{bin/malloc-comparison.csv}
  }
  \begin{tabnote}
    \Note{Note:}{Memory mapped addresses starting with the 0x2aa... prefix is an artifact of using \emph{make} to generate results. Executing the test program from \emph{bash} directly result in 0x7fff... prefixes.
    The difference is not important to our discussion.}
  \end{tabnote}
\end{table}

Table \ref{tab:mallocompare} illustrates how using a different allocator can affect potential aliasing between heap pointers.\footnote{Switching allocator was done by setting the LD\_PRELOAD environment variable.}
We observe the addresses of two equally sized \texttt{char} buffers allocated with \texttt{malloc} for different size parameters.
Equal three digit address suffix indicate an aliasing pair.
In addition to glibc, where the heap allocator is called \emph{ptmalloc}, we look at the following alternatives:
\begin{enumerate}
  \item Thread-Caching Malloc (tcmalloc) by Google~\cite{TCMalloc}.
  \item jemalloc, originally developed for FreeBSD~\cite{JEMalloc}.
  \item Hoard~\cite{Berger:2000:Hoard}.
\end{enumerate}
All of the above focus heavily on performance in multithreaded environments.
Heap allocation is intrinsically inefficient in that all threads share the same address space, leading to a high potential lock contention on memory accesses.
Here, we only look at behavior for a single thread, and whether the addresses returned alias or not.
